Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-12 paper.tex
arrumacoes de ordenacao e seccionamento das tabelas

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, \vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}. \opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations. This is important to coalesce with interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, academic groups,  \ngo s, and non-organized civil society.
    \item Derived \uri s, when reached via \http, can be redirected to where current documentation is held, as it is hosted by \url{http://purl.org}.
\end{itemize}

@@

\begin{table*}[!h]
    \footnotesize
  \centering
  \caption{Classes of the \ops\ (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{fig:owlCC}, these classes are related by the properties in Table~\ref{ospProps}. Noteworthy is \vcps\ class name ``{\tt vcps:Espa\c{c}o de A\c{c}\~ao}'', which caused original \vcps\ \owl\ to be pointed as corrupted by \protege\ (spaces are not allowed in \uri s). Also, class {\tt vcps:Role} was dropped as it yields logical problens in \vcps.}
  \begin{tabular}{|p{1.8cm}|p{1.6cm}||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}||p{3cm}|}\hline
      {\bf OPS class name} & {\bf VCPS class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} & {\bf upper ontology classes} \\\hline\hline
      Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) & {\tt bfo:'Material Entity'}, {\tt foaf:Person} \\ \hline
      Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) & {\tt bfo:'Material Entity'}, [\tt foaf:Organization] \\ \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results & {\tt bfo:'Material Entity'} \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly & {\tt bfo:'Material Entity'} \\ \hline
      Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) & {\tt bfo:'Material Entity'} \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a participatory role & {\tt bfo:'Material Entity'} \\ \hline
Participation-Characteristic & NivelDe-Participacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of the specific actor is happening & {\tt bfo:'Dependent Continuant'} \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for Action & {\tt bfo:'Dependent Continuant'} \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by Action & {\tt bfo:'Independent Continuant'} \\\hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation & {\tt bfo:'Processual Entity'} \\\hline
      Scope &{\bf Espa\c{c}o de A\c{c}\~ao} & Escopo & Ambito & Scope & the scope os Action & {\tt bfo:'Dependent Continuant'} \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action & {\tt bfo:'Dependent Continuant'} \\\hline
      Solution & Solucao & Solu\c{c}\~ao & Soluci\'on & Solution & solution achieved with Action & {\tt bfo:'Dependent Continuant'} \\\hline
      Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve & {\tt bfo:'Independent Continuant'} \\\hline
      {\color{red} dropped class}& Papel & -//-  & -//- & -//- & the role of the actor & -//- \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{table*}

\begin{table*}[!h]
  \centering
  \caption{Properties of the \ops\ (Ontology of Social Participation) along original \vcps\ names. Only a few ranges were established, and no domain, as these were not currently useful and, without them, \ops\ can be used more freely.}
  \begin{tabular}{|l|l||p{2.2cm}|p{2.2cm}|p{1.8cm}||l|p{2.2cm}|}\hline
{\bf OSP property name} & {\bf CVSP property name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf domain} & {\bf range} \\\hline\hline
theme & possuiTemaAssociado & tema & tema & theme & -//- & Theme \\ \hline
belongsTo & pertenceAoEspaco & pertence ao & pertence al & belongs to & -//- & Scope \\ \hline
supports & apoiaCausa & apoia & apoya & supports & -//- & -//- \\ \hline
contributesTo & compoeSolucao & contribui para & contribuye para & contributes to & -//- & -//- \\ \hline
executes & executaAcao & executa & ejecuta & executes & -//- & -//- \\ \hline
generates & geraCausa & gera & genera & generates & -//- & -//- \\ \hline
starts & iniciaCausa & inicia & inicializa & starts & -//- & -//- \\ \hline
solves & soluciona & soluciona & resuelve & solves & -//- & -//- \\ \hline
action & possuiAcao & a\c{c}\~ao & acci\'on & action & -//- & Action \\\hline
produces & produzResultado & produz & produce & produces & -//- & -//- \\\hline
proposes & propoeSolucao & prop\~oe & propone & proposes & -//- & -//- \\\hline
{\color{red} dropped} & temPapel & -//-  & -//-  & -//-  & -//-  & -//- \\ \hline
      trait & temNivelDeParticipacao & tra\c{c}o & rasgo & trait & -//- & -//- \\\hline
  \end{tabular}
  \label{ospProps}
\end{table*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (\ops). This image was rendered inside \protege, with the \owl\ code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} is more informative, but this diagram is more standard and might be simpler for the newcomer. Note that the taxonomic tree does not present any information about properties further linking these classes. The prefixes {\tt snap:} and {\tt span:} are \bfo\ specific.}
    \label{fig:owlCC}
\end{figure*}

@@

\subsection{Real use cases}\label{sec:real}
\opsi\ is a recent ontology. Even so, some real use cases can be pointed, from which are most notable:
\begin{itemize}
    \item The \pnud/\onu\ consultant contract 2013/00056, project BRA/12/018, was profoundly influenced by \ops~\cite{pnud5}. Within written products are other participatory ontologies, such as \opa, \ocd, \ontologiaa\ and \obs, which relates directly to \ops. Also, some methods for analysing \ops\ related data and for resource recommendation were delivered. These developments where done by the computational physics researcher R. Fabbri (first author of the present article), in collaboration with other parties, specially the Brazilian General Secretariat of the Republic, University of Bras\'ilia researchers, and free software parties. Data from three participatory instances were triplified: \participa, \aan\ and \cidadedemocratica; all related to \ops. These linked data resources are available in \sparql\ endpoints and can be dereferenced, in a similar fashion as done for \ops. As these are all executed in research facilities, they might lack maintenance and should be kept by a dedicated team.
    \item Another \pnud/\onu\ contract was responsible for some advances in information technologies. A special case is a dedicated \ops\ expansion\cite{paulo6}.
    \item The Brazilian Federal Data Processing Service (\serpro) is triplifying \participa\ data (the Brazilian federal social participation portal). Is this process, they as making direct use of \ops~\cite{tripSerpro}.
    \item The linked data Brazilian community had some maturing related to \ops\ beyond the ontologies developed, \pnud/\onu\ documentation, \serpro\ triplification, and \vcps. Interested government, civil society and academic parties circulated \ops\ documentation for conceptual and technological goals~\cite{circulaOps1,circulaOps2}.
\end{itemize}

@@

\section{Original diagram of OPS}\label{ap:dia}
Although deprecated, Figure~\ref{fig:diaorig} is the basis of \vcps\ and \ops. It is displayed here for completeness of exposition and as support for discussions within this article. 
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figs/diagramaOriginal}
    \caption{Original \vcps diagram. Although there were substantial \owl\ modifications, the main difference in the diagram is that the class \texttt{Role} was removed, as a \texttt{Role} is not able to start, execute or support anything. That is done by the Social Actor, as exposed in Figure~\ref{fig:v1}.}
    \label{fig:diaorig}
\end{figure*}

@@

@@

      \bfo  & Basic Formal Ontology & an upper ontology supporting information retrieval, analysis and integration in scientific and other domains \\hline

>>>>

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, \vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}. \opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations. This is important to coalesce interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, academic groups,  \ngo s, and non-organized civil society.
    \item Derived \uri s, when reached via \http, can be redirected to where current documentation is held, as it is hosted by \url{http://purl.org}.
\end{itemize}

@@

\begin{table*}[!h]
    \footnotesize
  \centering
  \caption{Classes of the \ops\ (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{fig:owlCC}, these classes are related by the properties in Table~\ref{ospProps}. Noteworthy is \vcps\ class name ``{\tt vcps:Espa\c{c}o de A\c{c}\~ao}'', which caused original \vcps\ \owl\ to be pointed as corrupted by \protege\ (spaces are not allowed in \uri s). Also, class {\tt vcps:Role} was dropped as it yields logical problens in \vcps.}
  \begin{tabular}{|p{1.8cm}|p{1.6cm}||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}||p{3cm}|}\hline
      {\bf OPS class name} & {\bf VCPS class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} & {\bf upper ontology classes} \\\hline\hline
      Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) & {\tt bfo:'Material Entity'}, {\tt foaf:Person} \\ \hline
      Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) & {\tt bfo:'Material Entity'}, {\tt foaf:Organization} \\ \hline \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results & {\tt bfo:'Material Entity'} \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly & {\tt bfo:'Material Entity'} \\ \hline
      Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) & {\tt bfo:'Material Entity'} \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a participatory role & {\tt bfo:'Material Entity'} \\ \hline \hline
Participation-Characteristic & NivelDe-Participacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of the specific actor is happening & {\tt bfo:'Dependent Continuant'} \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for Action & {\tt bfo:'Dependent Continuant'} \\\hline
      Scope &{\bf Espa\c{c}o de A\c{c}\~ao} & Escopo & Ambito & Scope & the scope os Action & {\tt bfo:'Dependent Continuant'} \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action & {\tt bfo:'Dependent Continuant'} \\\hline
      Solution & Solucao & Solu\c{c}\~ao & Soluci\'on & Solution & solution achieved with Action & {\tt bfo:'Dependent Continuant'} \\\hline \hline
      Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve & {\tt bfo:'Independent Continuant'} \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by Action & {\tt bfo:'Independent Continuant'} \\\hline \hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation & {\tt bfo:'Processual Entity'} \\\hline\hline
      {\color{red} dropped class}& Papel & -//-  & -//- & -//- & the role of the actor & -//- \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{table*}

\begin{table*}[!h]
  \centering
  \caption{Properties of the \ops\ (Ontology of Social Participation) along original \vcps\ names. Only a few ranges were established, and no domain, as these were not currently useful and, without them, \ops\ can be used more freely.}
  \begin{tabular}{|l|l||p{2.2cm}|p{2.2cm}|p{1.8cm}||l|p{2.2cm}|}\hline
{\bf OSP property name} & {\bf CVSP property name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf domain} & {\bf range} \\\hline\hline
theme & possuiTemaAssociado & tema & tema & theme & -//- & Theme \\ \hline
belongsTo & pertenceAoEspaco & pertence ao & pertence al & belongs to & -//- & Scope \\ \hline
action & possuiAcao & a\c{c}\~ao & acci\'on & action & -//- & Action \\\hline\hline
supports & apoiaCausa & apoia & apoya & supports & -//- & -//- \\ \hline
contributesTo & compoeSolucao & contribui para & contribuye para & contributes to & -//- & -//- \\ \hline
executes & executaAcao & executa & ejecuta & executes & -//- & -//- \\ \hline
generates & geraCausa & gera & genera & generates & -//- & -//- \\ \hline
starts & iniciaCausa & inicia & inicializa & starts & -//- & -//- \\ \hline
solves & soluciona & soluciona & resuelve & solves & -//- & -//- \\ \hline
produces & produzResultado & produz & produce & produces & -//- & -//- \\\hline
proposes & propoeSolucao & prop\~oe & propone & proposes & -//- & -//- \\\hline
      trait & temNivelDeParticipacao & tra\c{c}o & rasgo & trait & -//- & -//- \\\hline\hline
{\color{red} dropped} & temPapel & -//-  & -//-  & -//-  & -//-  & -//- \\ \hline
  \end{tabular}
  \label{ospProps}
\end{table*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (\ops). This image was rendered inside \protege, with the \owl\ code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} is more informative, but this diagram is more standard and might be simpler for the newcomer. Note that the taxonomic tree does not present any information about properties further linking these classes. The prefixes {\tt snap:} and {\tt span:} are \bfo\ specific for {\tt snap}shot and {\tt span}ning-time.}
    \label{fig:owlCC}
\end{figure*}

@@

\subsection{Real use cases}\label{sec:real}
\opsi\ is a recent ontology. Even so, some real use cases can be pointed, from which are most notable:
\begin{itemize}
    \item The \pnud/\onu\ consultant contract 2013/00056, project BRA/12/018, was profoundly influenced by \ops~\cite{pnud5}. Within written products are other participatory ontologies, such as \opa, \ocd, \ontologiaa\ and \obs, which relates directly to \ops. Also, some methods for analysing \ops\ related data and for resource recommendation were delivered. These developments were done by the computational physics researcher R. Fabbri (first author of the present article), in collaboration with other parties, specially the Brazilian General Secretariat of the Republic, University of Bras\'ilia researchers, and free software parties. Data from three participatory instances were triplified: \participa, \aan\ and \cidadedemocratica; all related to \ops. These linked data resources are available in \sparql\ endpoints and can be dereferenced, in a similar fashion as done for \ops. As these are all executed in research facilities, they might lack maintenance and should be kept by a dedicated team.
    \item Another \pnud/\onu\ contract was responsible for some advances in information technologies. A special case is a dedicated \ops\ expansion\cite{paulo6}.
    \item The Brazilian Federal Data Processing Service (\serpro) is triplifying \participa\ data (the Brazilian federal social participation portal). Is this process, they as making direct use of \ops~\cite{tripSerpro}.
    \item The linked data Brazilian community had some maturing related to \ops\ beyond the ontologies developed, \pnud/\onu\ documentation, \serpro\ triplification, and \vcps. Interested government, civil society and academic parties circulated \ops\ documentation for conceptual and technological goals~\cite{circulaOps1,circulaOps2}.
\end{itemize}

@@

\section{Original diagram of OPS}\label{ap:dia}
Although deprecated, Figure~\ref{fig:diaorig} is the basis of \vcps\ and \ops. It is displayed here for completeness of exposition and as support for discussions within this article. 
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figs/diagramaOriginal}
    \caption{Original \vcps\ diagram. Although there were substantial \owl\ modifications, the main difference in the diagram is that the class \texttt{Role} was removed, as a \texttt{Role} is not able to start, execute or support anything. That is done by the Social Actor, as exposed in Figure~\ref{fig:v1}.}
    \label{fig:diaorig}
\end{figure*}

@@

      \bfo  & Basic Formal Ontology & an upper ontology supporting information retrieval, analysis and integration in scientific and other domains \\\hline

@@
